---
title: "Reproducible figures and open science assignment"
header-includes: 
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output:
  pdf_document: default
  html_document: default
  word_document: default
  extra_dependencies: ["float"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff=50))
#setting the width of the pdf
```

### Loading all required packages-
```{r Data Exploration, message=FALSE, warning=FALSE, include=TRUE, results='hide'}
#install.packages(c("ggplot2", "palmerpenguins", "dplyr", "janitor", 
                   "svglite", "ggpubr", "ggstatsplot"))


library(palmerpenguins) #to load the dataset
library(dplyr) #to manipulate the data 
library(janitor) #to clean the dataset
library(ggplot2) #to make graphs
library(ggpubr) #to make the density plot
library(ggstatsplot) #to make the results figure
library(svglite) #to save the figures 
```

### Loading all function definitions:
```{r}
source("Functions/CleaningFunctions.r")
source("Functions/ggplotFunctions.r")
source("Functions/SavingFunctions.r")
```


```{r, echo=FALSE, warning=FALSE, results='hide'}
write.csv(penguins_raw, "penguins_raw.csv") #saving the dataset from the palmerpenguins package 
penguins_raw <- read.csv("penguins_raw.csv") #loading this dataset

bad_figure <- plot_bad_figure(penguins_raw) 
#producing the bad figure plot using a function from the ggplotFunctions script 

save_bad_figure_svg(penguins_raw,
                    "Figures/bad_figure.svg", 
                    size= 15, scaling=1) 
#saving this figure using a function from the SavingFunctions script
```

$\text{\underline{QUESTION 1: Data Visualisation for Science Communication}}$

*Create a figure using the Palmer Penguin dataset that is correct but badly communicates the data. **Do not make a boxplot**.*

### a) Provide your figure here:
```{r, echo=FALSE, warning=FALSE, fig.cap= " a graph showing flipper lengths of different sexes, exemplifying poor graph design"}
bad_figure
```


### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

'A scientific figure is, by definition, tied to the data' (Rougier, *et al*. 2014) therefore, figures must convey an objective, accurate representation of the data. The lack of jitter on the above graph obscures individual data points. As such, much of the raw data is absent making it appear as if the sample size is much smaller than it is and in turn, the power. Next, the limits of the y axis are inappropriate as the minimum data point is at 172mm, not near 0mm. The data is then compressed to fit on this inappropriately long axis, obscuring the true distribution of the data as the graph makes it seem as if the range of the data is smaller than in reality. An incorrect choice of scale is a common error in many scientific graphs (Franzblau and Chung 2012) The data obfuscation caused by the lack of jitter and inappropriate y axis is further enhanced by the inappropriate colour scales which make the different islands indistinguishable. Next, not including a title means the graph is not in context. The only information about the animals in the study are that they have flippers, but this is still taxonomically diverse so, readers may apply to the data to the wrong animal- a title would make the sample used clear. Finally, the 'NA' are not technically datapoints, just a string of flipper lengths so should not be graphically represented. No statistical analysis can include these values and as such, their inclusion may deter readers away from the actual datapoints. Overall, the above graph contains obscured data and points that are technically not data, neither of which are in context therefore, the true underlying data and its relevance are not indicated. 

* Franzblau, L.E. and Chung, K.C., (2012), 'Graphs, Tables, and Figures in Scientific Publications: The Good, the Bad, and How Not to Be the Latter', *The Journal of Hand Surgery*, 37(3), pp. 591-596. doi: https://doi.org/10.1016/j.jhsa.2011.12.041 
* Rougier, N.P, *et al*., (2014), 'Ten Simple Rules for Better Figures', *PLOS Computational Biology*, 10(9), pp.e1003833. doi: https://doi.org/10.1371/journal.pcbi.1003833 

$\text{\underline{QUESTION 2: Data Pipeline}}$

### *Introduction:*
Sexual dimporphism is an established phenomenon in the animal kingdom (McPherson and Chenoweth 2012). The overarching theme regarding evolution is increasing reproductive success, but in the case of sexual dimorphism, this success tends to be restricted to males where variation differentially affects the chances of that male mating. I will be investigating sexual dimorphism in flipper length in *Pygoscelis* penguins, inspired by research conducted by Gorman *et al.* (2014) in the Palmer archipelago of Antarctica. Notably, the palmerpenguins package is based off their data. 

* Gorman, K.B. *et al*., (2014), 'Ecological Sexual Dimorphism and Environmental Variability within a Community of Antarctic Penguins (Genus *Pygoscelis*)', *PLOS ONE*, 9(3), pp. e90081. doi: https://doi.org/10.1371/journal.pone.0090081 
* McPherson, F.J. and Chenoweth, P.J., (2012), 'Mammalian sexual dimorphism', *Animal Reproductive Science*, 131(3-4), pp. 109-122. doi: https://doi.org/10.1016/j.anireprosci.2012.02.007 

### Loading and cleaning penguin data-
```{r}
write.csv(penguins_raw, "PenguinData/penguins_raw.csv") 
#saving the 'penguins_raw' data from the palmerpenguins package in the 'PenguinData' folder 
penguins_raw <- read.csv("PenguinData/penguins_raw.csv")
#loading the data from the project folder
```

```{r}
names(penguins_raw) #checking the raw data column titles
```

```{r, results='hide'}
penguins_clean <- penguins_raw %>%
    clean_function() %>% 
    clean_column_names() %>% 
    shorten_species() %>% 
    remove_empty_columns_rows() 
#cleaning the raw data using functions from the CleaningFunctions script
```

```{r}
names(penguins_clean) #check that the functions have worked 
```

```{r}
write.csv(penguins_clean, "PenguinData/penguins_clean.csv")
#saving the "penguins_clean' data in the 'PenguinData' folder
```


### *Hypothesis:*
Now that the data has been cleaned, we can produce figures and conduct statistical tests to test sexual dimorphism in flipper length.

Preliminary hypotheses:

* H0- There is no significant difference in mean flipper length between male and female Adelie, Gentoo, and Chinstrap penguins

* H1- There is a significant difference in mean flipper length between male and female Adelie, Gentoo, and Chinstrap penguins

### Raw data showing sex and flipper length-
```{r, results='hide', fig.cap= "a plot showing the flipper length of penguins, separated by sex"}
#plotting the explanatory figure using functions from the ggplotFunctions script
explanatory_figure <- plot_explanatory_figure(penguins_clean)
explanatory_figure

#saving the figure using a function in the SavingFunctions script
save_explanatory_plot_svg(penguins_clean, 
                          "Figures/explanatory_figure.svg",
                          size=15, scaling=1)
```

### *Statistical Methods:*
Testing the assumptions of a two-sample t test (for the purposes of this assignment, I am going to assume that the sample is random and independent)

### Testing that the data is normally distributed:
### Visual summary of the data distribution-
```{r, warning=FALSE, message=FALSE, fig.cap= "density plot of flipper lengths indicating a lack of normal distribution"}
#plotting a density graph using a function from the ggplotFunctions script 
density_plot <- plot_density_plot(penguins_clean)
density_plot
```

### Statistical summary of the data distribution-
```{r}
shapiro.test(penguins_clean$flipper_length_mm) #goodness of fit of a normal distribution to the data 
```
* H0- the data is normally distributed
* H1- the data is not normally distributed 
* p<0.05 therefore, we must reject H0, the data deviates significantly from a normal distribution 

### Testing for homoscedasticity- 

* the test for homoscedasticity requires a normal distribution so, we cannot conduct this test 

**Overall, the data does not meet the assumtpions of a two-sample t-test so, we must use the non-parametric version, the Mann-Whitney U test**

### Mann-Whitney U test:

Final hypotheses:

* H0- flipper lengths are the same between male and female *Pygoscelis* penguins
* H1- flipper lengths are different between male and female *Pygoscelis* penguins

```{r}
wilcox.test(flipper_length_mm ~ sex, data = penguins_clean) #Mann-Whitney U test
```
p<0.05 therefore, reject H0, the flipper lengths are significantly different between male and female *Pygoscelis* penguins indicating sexual dimorphism in flipper length 


### *Results and Discussion:*

### Results figure-
```{r, results='hide', fig.cap="results figure of the flipper length Mann-Whitney U test"}
#plotting the results figure using functions from the ggplotFunctions script 
results_figure <- plot_results_figure(penguins_clean)
results_figure

#saving the figure using a function in the SavingFunctions script
save_results_plot_svg(penguins_clean, 
                      "Figures/results_figure.svg", 
                      size=15, scaling=1)
```

As shown in figure 4, female and male flipper length is significantly different in *Pygoscelis* penguins. For example, the median flipper length is 193mm in females and 200.5mm in males. Larger male phenotypes are common in sexually dimorphic birds, so it is not surprising that my analysis also found this pattern. Speculations of the selective pressures favouring larger male flipper lengths in the literature are outweighed by explanations of the consequences of this dimorphism. For example, Gorman *et al.* (2014) found that pre-breeding trophic niches have diverged between males and females as a result of sexual size dimorphism. Notably, there is a glaring gap in the literature of the influence of sexually dimoprhic flipper lengths in *Pygoscelis* penguins, with many analyses finding a lack of sexual dimorphism in flipper length. It is therefore of crucial significance to study flipper lengths in *Pygoscelis* penguins to fully understand the role of male and female phenotypes in niche differences. 

### *Conclusion:*

In this analysis, I have found that *Pygoscelis* penguins, including those of the Adelie, Gentoo and Chinstrap species, exhibit sexual dimorphism in flipper length. Interestingly, previous work has not revealed flipper length as a sex determinant in these penguins (Gorman *et al.* 2014). While it would be worthwhile to study flipper length dimorhpism further, confirming the reliability of this current work, I have shown potential sex differences of flipper length in *Pygoscelis* penguins. Additionally, allospecific differences in sexual dimorphism would be interesting to investigate further, to assess whether the differing ecological niches of the three *Pygoscelis* species are evident in their degree of sexual dimoprhism. Overall, my analysis reveals the potential of flipper length as a sexually dimorphic phenotype in *Pygoscelis* species, with the reasons of such being an interesting topic to investigate further. 

------------------------------------------------------------------------

